A Classical Density-Functional Theory for Describing Water Interfaces

### 2011-10-19 paper/paper.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory and Methods}
We introduce a new classical density functional for water that
\needsworklater{predicts the vapor density, liquid density, bulk
  modulus, and bulk surface tension, as well as allows for
  liquid-vapor coexistence}.  The Helmholtz free energy functional is
based on SAFT, and is composed of the usual terms:
\begin{equation}
  F[n] = F_\textit{id}[n] + F_\textit{hs}[n] + F_\textit{assoc}[n] + F_\textit{disp},
\end{equation}
where $F_\textit{id}$ is the ideal gas free energy, $F_\textit{hs}$ is
a hard-sphere free energy, $F_\textit{assoc}$ is the free energy of
association, and $F_\textit{disp}$ is the dispersion energy.  In
addition, a chemical potential term is used, so we actually work in
the grand canonical ensemble.  In the following sections, we will
introduce the terms of this functional.

\subsection{Ideal gas functional}
The first term is the ideal gas free energy functional,
which incorporates effects of kinetic energy.  The ideal gas
functional is given by
\begin{equation}\label{idealgas}
  F_{id}[n] = k_B T \int n(\xx)\left( \ln{n(\xx)} - 1\right) d\xx
\end{equation}
where n(\xx) is the density of water molecules.  The ideal gas free
energy functional has the property that the contact density at a hard
surface $n_c$ is proportional to the pressure on that surface,
according to the equation
\begin{equation}
  p = n_c k_BT \:,
\end{equation}
which also leads to the property that the free energy of solvation of
a hard solute is proportional to its volume in the limit of small
solutes.  These properties are retained by the total functional,
provided the remaining terms are \emph{purely
  nonlocal}~\cite{ashcroft?}.

\subsection{Hard-sphere repulsion}
The repulsive intermolecular interactions are modeled using a
Fundamental\nobreakdash-Measure Theory~\cite{rosenfeld1997}(FMT) functional,
which combines scaled-particle theory and the Percus-Yevick theory,
and an attractive interaction.  The FMT functional treats
each water molecule as a hard sphere with a radius approximately that of
a water molecule $(R~=~2.3~bohr)$.\\

We model the repulsive interactions using the White Bear version of
the Fundamental-Measure Theory~(FMT) functional published in
reference~\cite{roth2002whitebear}.  \needsworklater{The FMT
  functional describes the excess free energy of a hard-sphere
  fluid. Fundamental-Measure Theory has several advantages over other
  approaches, one of the most important being the simplicity of
  calculations.  This particular FMT also reliably predicts bulk
  properties of fluids in one, two and three dimensions, as well as
  exactly determines the excess free energy in the zero dimensional
  limit.}
\begin{equation}
F_{hs}[n] = k_B T \int (\Phi_1(\xx) + \Phi_2(\xx) + \Phi_3(\xx)) d\xx \; ,
\end{equation}
with integrands
\begin{align}
\Phi_1 &= -n_0 \ln\left( 1 - n_3\right)\\
\Phi_2 &= \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3} \\
\Phi_3 &= \frac{
  \frac13 n_2^3 - n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}
}{
  8\pi \left( 1 - n_3 \right)^2
} ,
\end{align}
using the weighted densities
\begin{align}
  n_3(\xx) &= \int n(\xx') \Theta(\left|\xx - \xx'\right| - R) d\xx' \\
  n_2(\xx) &= \int n(\xx') \delta(\left|\xx - \xx'\right| - R) d\xx'
  \\
  \mathbf{n}_{V2} &= \mathbf{\nabla} n_3 \\
  n_1 &= \frac{n_2}{4\pi R}\\
  \mathbf{n}_{V1} &= \frac{\mathbf{n}_{V2}}{4\pi R}\\
  n_0 &= \frac{n_2}{4\pi R^2}
\end{align}

@@

\subsection{Association free energy}
To the purely repulsive hard-sphere free energy, we add two attractive
energy terms.  The first is the free energy of \emph{association},
which describes the effects of hydrogen bonds.  These are modeled as
four attractive patches (``association sites'') on the surface of the
hard sphere.  These four sites represent the two hydrogens and two
electron lone pairs.  There is an attractive energy
$\epsilon_\textit{assoc}$ when two molecules are arranged such that
the hydrogen of one interacts with the lone pair of the other.  The
volume of this interaction is $\kappa_\textit{assoc}$.  Thus the
association term in the free energy has two free parameters that must
be fit based on experimental data.

The association free energy has the form
\begin{align}
  F_\text{assoc}[n] &= 4 k_BT \int n_0(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx
\end{align}
where the value of $4$ comes from the four association sites per
molecule, and the functional $X$ is the fraction of assotiation sites
\emph{not} hydrogen-bonded.  The fraction $X$ is determined by the
quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 8n_0(\xx)\zeta(\xx)\Delta(\xx)} - 1}
  {4 n_0(\xx)\zeta(\xx)\Delta(\xx)}
\end{align}
where $\zeta$ comes from
reference\cite{yu2002fmt-dft-inhomogeneous-associating,
  fu2005vapor-liquid-dft} and describes the inhomogeneity of the fluid
density:
\begin{align}
  \zeta(\xx) &= 1 - \frac{\mathbf{n_{2V}}\cdot\mathbf{n_{2V}}}{n_2^2}
\end{align}
The function $\Delta$ is given... below?

The following contact density for the hard-sphere fluid is in the code
called \texttt{gHS}, and comes
from\cite{yu2002fmt-dft-inhomogeneous-associating,
  fu2005vapor-liquid-dft}, but I don't use it.
\begin{align}
  \Delta &= 4\pi \kappa_\textit{assoc} g^\textit{HS}(\sigma)e^{-\beta
    \epsilon_\textit{assoc}} \\
  g^\textit{HS}(\sigma) &= \frac1{1-n_3}\left(1+\frac14\frac{\zeta n_2}{1-n_3}
  \left(1 + \frac{n_2}{14 (1-n_3)}\right)\right)
\end{align}
\textcolor{red}{Here is paper where I got the association functional:}

Instead, I use
\begin{align}
  \Delta(\xx) &= \kappa_\textit{assoc}
  e^{-\beta\epsilon_\textit{assoc}} g^\textit{SW}_\sigma(\xx) \\
  g^\textit{SW}_\sigma(\xx) &= g^\textit{HS}_\sigma(\xx) +
  \frac{1}{4}\beta\left(\frac{\partial a_1(\xx)}{\partial n_3(\xx)} -
  \frac{\lambda}{3 n3(\xx)}\frac{\partial a_1(\xx)}{\partial \lambda}\right)
\end{align}
% This is equation 77 of gil-villegas-1997-SAFT-VR
which I got from reference\cite{gil-villegas-1997-SAFT-VR}.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/saturated-liquid-density}
\end{center}
\caption{The theoretical versus experimental saturated liquid density
  versus temperature.  }
\label{fig:saturated-liquid-density}
\end{figure}

\subsection{Dispersion free energy}
The final term in the free energy is the dispersion term.  We use a
dispersion term based on the SAFT-VR
approach\cite{gil-villegas-1997-SAFT-VR}, which has two free
parameters, an interaction energy $\epsilon_\textit{disp}$ and a
length scale $\lambda_\textit{disp}$.

The dispersion free energy has the form~\cite{gil-villegas-1997-SAFT-VR}
\begin{align}
  F_\text{disp} &= A_1 + \beta A_2
\end{align}
where $A_1$ and $A_2$ are the first two terms in a high-temperature
perturbation expansion.  $A_1$ is the mean-field dispersion
interaction, which for a homogeneous fluid is given by
\begin{align}
  A_1 &= \frac12 n_b \int \varphi(\left|\xx\right|)
  g_{HS}(\left|\xx\right|) d\xx
\end{align}
The second dispersion term in the free energy $A_2$ describes the
effect of fluctuations resulting from the compression of the fluid due
to the dispersion interaction itself, and is commonly approximated
using the local compressibility approximation (LCA).

We use a free square-well function for the dispersion $\varphi$, which
has been demonstrated \textcolor{red}{(CITE)} to give a reasonable fit
to the equation of state.  Thus our model interaction has the form:
\begin{equation}
  \varphi(r) = \Theta(r-2 \lambda_\textit{disp} R)
\end{equation}
where $\Theta$ is the Heaviside step function.

\textcolor{red}{TODO:  Add the actual functions that we use and the
  actual form.  The above is really just an introduction.}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/temperature-versus-density}
\end{center}
\caption{The theoretical versus experimental saturated liquid density
  versus temperature.  }
\label{fig:saturated-liquid-density}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/entropy}
\end{center}
\caption{The free energy, internal energy, and temperature*entropy versus density at room temperature.  }
\label{fig:energy-room-temp}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/entropy-at-690K}
\end{center}
\caption{The free energy, internal energy, and temperature*entropy versus density near the critical temperature.  }
\label{fig:energy-near-critical-temp}
\end{figure}

From the Helmholtz free energy functional, we may obtain any other
thermodynamic functions we should choose.  The grand free energy
$\Phi$, which we ordinarily use in our computations, is obtained by
simply adding a chemical potential term:
\begin{equation}
  \Phi[n] = F[n] + \mu \int n(\xx) d\xx
\end{equation}

The pressure would naturally be obtained by taking a partial
derivative with respect to volume at fixed temperature.  Since
\emph{volume} isn't a parameter in density-functional theory, we
consider the \emph{free energy per unit volume}.
\begin{align}
  F[n] &= f[n]V \\
  p[n] &= -\left(\frac{\partial F[n]}{\partial V}\right)_{T} \\
  &= -\left(\frac{\partial f[n]V}{\partial V}\right)_{T} \\
  &= -V\left(\frac{\partial f[n]}{\partial V}\right)_{T}
   - f[n]\left(\frac{\partial V}{\partial V}\right)_{T} \\
  &= n \left(\frac{\partial f[n]}{\partial n}\right)_{T} - f[n]
\end{align}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/pressure-with-isotherms}
\end{center}
\caption{The pressure versus density for various temperatures, including experimental pressure data from NIST.  }
\label{fig:pressure-with-isotherms}
\end{figure}

@@

Once we have the entropy, finding the internal energy is a simple
matter of subtraction:
\begin{align}
  U[n] &= F[n] + TS[n]
\end{align}
and the enthalpy is not much more work:
\begin{align}
  H[n] &= F[n] + TS[n] - pV
\end{align}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/surface-tension}
\end{center}
\caption{The theoretical versus experimental surface tension
  versus temperature.  }
\label{fig:surface-tension}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/surface-298}
\end{center}
\caption{Liquid-vapor interface profile at room temperature.  }
\label{fig:liquid-vapor-profile}
\end{figure}

\subsection{Crossover formalism for handling critical behavior}

I am currently working on implementing the crossover approach of
Kiselev and Ely which is designed to make the critical behavior come
out right\cite{kiselev2006new}.  This method works by defining a
\emph{classical} free energy $f_0$---which where classical really is
closer to \emph{mean field} in meaning, and indicates a theory that
doesn't include long-range fluctuations.  This free energy is then
used to construct a corrected free energy by feeding portions of it a
``fake'' density and temperature that are constructed so as to
reconstruct the correct universal scaling laws near the critical
point.

@@

In order to ensure that the empirical parameters are correct, we test
the vapor pressure (see Figure~\ref{fig:equation-of-state}) and
saturated liquid density (see
Figure~\ref{fig:saturated-liquid-density}) against experimental
values.  To find these densities, we use the ``common tangent'' rule,
as illustrated in Figure~\ref{fig:homogeneous}.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/finding-vapor-pressure}\\
\includegraphics[width=\columnwidth]{figs/near-critical-point}
\end{center}
\caption{By plotting the free energy vs. density it is possible to
  determine a saturated liquid or vapor state by the common tangent
  method.  }
\label{fig:homogeneous}
\end{figure}

\begin{table}
\begin{tabular}{ccc}
  Fitting Parameters& \hspace{2em} & Experimental Values\\
\begin{tabular}{cc}
  $R$ & 2.3 $a_o$ \\
  $A$ & -0.0100909 Hartree \\
  $\gamma$ & 5\\
  $V$ & 515.792 $a_o^3$  \\
  $b$ & 0.75 $a_o$ \\
  $\mu$ & -0.014063 Hartree 
\end{tabular}
&& 
\begin{tabular}{cc}
 $n_l$ & 4.939 $\times 10^{-3} a_o^{-3}$\\
 $n_v$ & 1.141 $\times 10^{-7} a_o^{-3}$\\
 $\sigma$ & ??? \\
 $B$ & $7.435\times 10^{-5}$ Hartree$/a_o^3$\\
 $T$ & 298.15 K \\
\end{tabular}
\end{tabular}

\caption{Empirical parameters of optimized functional and experimental
  values matched, where $a_o$ is the Bohr radius, and $\sigma$ is the
  surface tension.  This table is not yet complete, but it should
  be soon...  We'll also need to think about the best units to use
  here.  Atom units might be best, but it also might be better to
  consistently use eV and angstrom or something.  I'm not sure about
  putting $k_BT$ data in the table.  Maybe it belongs in the caption?}
\label{tab:parameters}
\end{table}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

The simple yet complex water molecule is one of the most important 
substances for life and industry. The structure and geometry of the 
water molecule leads to a strong polarity and complex intermolecular
forces. The hydrogen bonds formed by water are very strong, creating
tetragonal clusters that lead to many of water's unusual properties,
such as the high melting, boiling, and critical temperatures as well as the
density maximum above the melting point.
The large dielectric constant is partially responsible for
water acting as a universal solvent, along with the ease of availability and
use.
There are many models used to describe water since so many different
processes occur in aqueous solution. It is challenging to describe all the 
relevant properties of water at once, especially if a low computational cost is
also desired.

Some models are primarily designed for use with homogeneous systems. Although
these 
models are useful in predicting some bulk properties, the interesting
chemistry that occurs in water cannot be described with a homogeneous-only
model. To
describe solvation and other processes in aqueous solution, a more robust theory
that 
can handle varying densities is needed.

Classical density-functional theory (DFT) provides an efficient way to evaluate
the free energy using the probability distribution of the fluid. The free energy 
is a functional
of the density function and is independent of any external
potentials\cite{ebner1976density}. The most common use of DFT is calculating ground
state electron densities in quantum mechanical systems. The Hohenberg-Kohn
theorem\cite{hohenberg1964inhomogeneous} allows us to write the ground state
energy in terms of a universal functional of the electron density and a separate
term including the external potential. Mermin\cite{mermin1965thermal} gives a
theorem extending that of Hohenberg and Kohn to nonzero temperatures:
\begin{equation}
  A(T) = \underset{n(r)}{min}\left\{ F[n(r),T] + \int V_{ext}(r) n(r)
d^3r\right\}
\end{equation}
This is the basis for classical DFT, where the relevant density is not the
electron density,
but the density of molecules. This can be used with a variety of 
external potentials, from the simple interaction with a 
hydrophobic wall to the more complex, stronger interaction with hydrophilic
solutes.

\subsection{Statistical Associating Fluid Theory}

The high hydrogen bonding and strong polar interactions in water characterize it
as an associating fluid, and thus it requires a different treatment than simple
fluids which are mainly defined by van der Waals forces and weak electrostatic
interations. The Statistical Associating Fluid Theory (SAFT) was designed for
use with fluids such as water, to account for the important association forces
(like hydrogen bonds) as well as any effects due to clusters or chains forming
in the liquid\cite{muller2001molecular}. It is based on Wertheim's first-order
thermodynamic perturbation theory (TPT1)
\cite{wertheim1984fluidsI,wertheim1984fluidsII,wertheim1986fluidsIII,
wertheim1986fluidsIV} which provides a useful way to account for strong
associative interations between molecules.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory and Methods}
We introduce a new classical density functional for water that
predicts the vapor pressure, vapor density, liquid density, bulk surface tension, 
and enthalpy, as well 
as allows for liquid-vapor coexistence.  The Helmholtz free energy functional is
based on SAFT, and is composed of the usual terms:
\begin{equation}
  F[n] = F_\textit{id}[n] + F_\textit{hs}[n]  +
F_\textit{disp}[n]+ F_\textit{assoc}[n],
\end{equation}
where $F_\textit{id}$ is the ideal gas free energy, $F_\textit{hs}$ is
the hard-sphere free energy, $F_\textit{disp}$ is the dispersion energy,
and $F_\textit{assoc}$ is the free energy of association.  In
addition, a chemical potential term is used, so we actually work in
the grand canonical ensemble.  In the following sections, we will
introduce the terms of this functional.

\subsection{Ideal gas functional}
The first term is the ideal gas free energy functional,
which incorporates effects of kinetic energy.  The ideal gas
functional is given by
\begin{equation}\label{idealgas}
  F_{id}[n] = k_B T \int n(\xx)\left( \ln{\frac{n(\xx)}{n_Q}} - 1\right) d\xx
\end{equation}
where n(\xx) is the density of water molecules and $n_Q$ is the quantum 
concentration. This is the particle concentration where quantum effects
become appreciable and is given by
\begin{equation}\label{quantumconcentration}
 n_Q =\left(\frac{mk_BT}{2\pi\hbar^2}\right)^{3/2}
\end{equation}
where $m$ is the molecular mass of water. The ideal gas free
energy functional has the property that the contact density at a hard
surface $n_c$ is proportional to the pressure on that surface,
according to the contact value theorem
\begin{equation}\label{contactvaluethm}
  p = n_c k_BT \:,
\end{equation}
which also leads to the property that the free energy of solvation of
a hard solute is proportional to its volume in the limit of small
solutes.  These properties are retained by the total functional,
provided the remaining terms are \emph{purely nonlocal}. This requires
that
\begin{equation}
 \frac{\delta^2F_{\textit{excess}}[n]}{\delta n(\xx_1)~\delta n(\xx_1)}\neq\infty
\end{equation}
where $F_{\textit{excess}}[n]$ describes all the remaining terms in the functional.
This ensures that any higher order terms in the Taylor expansion
of $F[n+\Delta n]$ are small and do not contribute to the small
solute free energy of solvation.

\subsection{Hard-sphere repulsion}

We model the repulsive interactions using the White Bear version of
the Fundamental-Measure Theory~(FMT) functional for the hard-sphere
fluid~\cite{roth2002whitebear}.  In the homogeneous limit, the White
Bear functional reduces to the Mansoori-Carnahan-Starling-Leland
equation of state.  FMT functionals are expressed as the integral of
the \emph{fundamental measures} of a fluid, which provide local
measures of quantities such as the filling fraction, density of
spheres touching a given point and mean curvature.  The excess free
energy is written as:
\begin{equation}
F_{hs}[n] = k_B T \int (\Phi_1(\xx) + \Phi_2(\xx) + \Phi_3(\xx)) d\xx \; ,
\end{equation}
with integrands
\begin{align}
\Phi_1 &= -n_0 \ln\left( 1 - n_3\right)\\
\Phi_2 &= \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3} \\
\Phi_3 &= (n_2^3 - 3n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2})
  \frac{
    n_3 + (1-n_3)^2\ln(1-n_3)
  }{
    36\pi n_3^2\left( 1 - n_3 \right)^2
  } ,
\end{align}
where the fundamental measure densities are given by:
\begin{align}
  n_3(\xx) &= \int n(\xx') \Theta(\left|\xx - \xx'\right| - R) d\xx' \\
  n_2(\xx) &= \int n(\xx') \delta(\left|\xx - \xx'\right| - R) d\xx'
  \\
  \mathbf{n}_{V2} &= \mathbf{\nabla} n_3 \\
  n_1 &= \frac{n_2}{4\pi R}\\
  \mathbf{n}_{V1} &= \frac{\mathbf{n}_{V2}}{4\pi R}\\
  n_0 &= \frac{n_2}{4\pi R^2}.
\end{align}
The density $n_3$ is the filling fraction and $n_2$ describes the number
of spheres touching a given point. For our water functional, we use the 
hard-sphere radius of
3.03420~\AA, which was found to be optimal by Clark
\emph{et al}.\cite{clark2006developing}

\newcommand\etadisp{\ensuremath{\eta_\textit{d}}}
\newcommand\epsilondisp{\ensuremath{\epsilon_\textit{d}}}
\newcommand\epsilonassoc{\ensuremath{\epsilon_\textit{a}}}
\newcommand\kappaassoc{\ensuremath{\kappa_\textit{a}}}
\newcommand\lambdadisp{\ensuremath{\lambda_\textit{d}}}
\newcommand\lscale{\ensuremath{s_d}}
\subsection{Dispersion free energy}
The first term in the attractive energy is the dispersion term.  This
will include van der Waals attraction and any orientation-independent
interactions. We use a
dispersion term based on the SAFT-VR
approach\cite{gil-villegas-1997-SAFT-VR}, which has two free
parameters, an interaction energy $\epsilondisp$ and a
length scale $\lambdadisp$.

The dispersion free energy has the form~\cite{gil-villegas-1997-SAFT-VR}
\begin{align}
  F_\text{disp}[n] &= \int (a_1(\xx) + \beta a_2(\xx))d\xx
\end{align}
where $a_1$ and $a_2$ are the first two terms in a high-temperature
perturbation expansion and $\beta=1/k_BT$.  The first term $a_1$ is 
the mean-field dispersion
interaction, which reduces in the homogeneous limit to
\begin{align}\label{eq:A1-simple}
  a_1 &= \frac12 n_b \int \varphi(\left|\xx\right|)
  g_{HS}(\left|\xx\right|) d\xx
\end{align}
where $n_b$ is the bulk density and $g_{HS}$ is the homogeneous
hard-sphere fluid correlation function.
The second dispersion term in the free energy $a_2$ describes the
effect of fluctuations resulting from the compression of the fluid due
to the dispersion interaction itself, and is commonly approximated
using the local compressibility approximation (LCA). In the LCA,
we assume the energy fluctuation is directly related to the
compressibility of a hard-sphere reference fluid\cite{barker1976liquid}.

We use a free square-well function for the dispersion $\varphi$, which
is the choice used by Clark \emph{et al}~\cite{clark2006developing},
and allows a reasonable fit to the equation of state.  Thus our model
interaction has the form:
\begin{equation}
  \varphi(r) = \Theta(r-2 \lambdadisp R)
\end{equation}
where $\Theta$ is the Heaviside step function.

The form of $a_1$ is given in
reference~\cite{gil-villegas-1997-SAFT-VR}, but is expressed in terms
of the filling fraction.  In order to apply this form to an
inhomogeneous density distribution, we construct an effective local
filling fraction for dispersion $\etadisp$, given by
\begin{align}
  \etadisp(\xx) &=
  \int \frac{\Theta(2\lambdadisp\lscale R - \left|\xx-\xx'\right|)}{
             \left(2\lambdadisp\lscale\right)^3} n(\xx') d\xx'.
\end{align}
This effective filling fraction is used throughout the dispersion
calculations, and represents a filling fraction averaged over the
effective range of the dispersive interaction.  This functional
introduces an additional empirical parameter $\lscale$ which adjusts
the length scale over which the dispersion interaction is correlated.

The first term in the dispersion functional $a_1$ when written using
the above filling fraction for dispersion has the form
\begin{align}
  a_1 &= 
   -4(\lambdadisp^3 - 1)\epsilondisp \etadisp(\xx)
    g^{HS}_\sigma(\eta_\textit{eff}(\xx)) \\
  g_\sigma^{HS}(\eta) &= \frac{1}{1-\eta}
  +\frac32\frac{\zeta(\xx)}{(1-\eta)^2}
  + \frac12\frac{\eta^2\zeta(\xx)}{(1-\eta)^3}
  \\
  \zeta(\xx) &= 1 - \frac{\mathbf{n}_{2V}\cdot\mathbf{n}_{2V}}{n_2^2}\\
  \eta_\textit{eff}(\xx) &= \sum_{ij=0}^3 C_{ij}  \etadisp^i(\xx)
  \lambdadisp^j
\end{align}
where $g_\sigma^{HS}$
is the hard-sphere fluid correlation function evaluated at contact.
The $\zeta$ factor
(also seen in the association energy, below) is a dimensionless
measure of the local density inhomogeneity. The $C_{ij}$ values
 are numerical constants taken from
reference\cite{gil-villegas-1997-SAFT-VR}, which come from a numerical
fit to the integral in Equation~\ref{eq:A1-simple} over a range of
values for $\lambdadisp$ and filling fractions.

The second term, $a_2$, which describes the contribution to the free
energy associated with fluctuations is given by
\begin{align}
  a_2 &= \frac12 \epsilondisp
              K_{HS}(\etadisp(\xx)) \etadisp(\xx)
              \frac{\partial a_1}{\partial \etadisp(\xx)}
\end{align}
where $K_{HS}$ is the Percus-Yevick hard-sphere compressibility, given
by
\begin{align}
  K_{HS} &=
    \frac{(1 - \etadisp)^4}{1 + 4(\etadisp + \etadisp^2)}.
\end{align}

\subsection{Association free energy}
The final attractive energy term is the association term, which
accounts for hydrogen bonding.  Hydrogen bonds are modeled as four
attractive patches (``association sites'') on the surface of the hard
sphere.  These four sites represent two protons and two electron lone
pairs.  There is an attractive energy $\epsilonassoc$ when
two molecules are oriented such that the proton of one overlaps
with the lone pair of the other.  The volume over which this
interaction occurs is $\kappaassoc$, giving the association
term in the free energy has two empirical parameters that are fit to
experimental data.

The association functional we use is a modified version of that of Yu
and Wu\cite{yu2002fmt-dft-inhomogeneous-associating}.\footnote{We
  should note that Fu and Wu\cite{fu2005vapor-liquid-dft} use almost
  the same functional, but their paper contains errors in the
  association term and is not useful as a reference.}  The association
functional of Fu and Wu includes the effects of density
inhomogeneities on the \emph{contact density} of hard spheres, which
conventionally appears in the form of the density multiplied by the
contact value of the correlation function $g^{HS}_\sigma$, but is
based on the SAFT-HS equation of state, rather than the SAFT-VR
equation of state\cite{gil-villegas-1997-SAFT-VR} which is used in the
optimal SAFT parametrization for water found by Clark \emph{et
  al}\cite{clark2006developing}.

The association free energy for our four-site model has the form
\begin{align}
  F_\text{assoc}[n] &= 4 k_BT \int n_0(\xx)\zeta(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx
\end{align}
where the value of $4$ comes from the four association sites per
molecule, and the functional $X$ is the fraction of assotiation sites
\emph{not} hydrogen-bonded, and $\zeta(\xx)$ is a dimensionless
measure of the density inhomogeneity.  The fraction $X$ is determined
by the quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 8n_0(\xx)\zeta(\xx)\Delta(\xx)} - 1}
  {4 n_0(\xx)\zeta(\xx)\Delta(\xx)}
\end{align}
where the factor $\zeta$ from
Yu and Wu\cite{yu2002fmt-dft-inhomogeneous-associating,
  fu2005vapor-liquid-dft} describes the inhomogeneity of the fluid
density, and reduces to unity in the homogeneous limit:
\begin{align}
  \zeta(\xx) &= 1 - \frac{\mathbf{n}_{2V}\cdot\mathbf{n}_{2V}}{n_2^2}.
\end{align}
The functional $\Delta$ which appears in the formula for $X$ is a
measure of hydrogen-bonding probability.  In the SAFT-VR
model\cite{gil-villegas-1997-SAFT-VR}, $\Delta$ is given by
\begin{align}
  \Delta(\xx) &= \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right) \\
  g^\textit{SW}_\sigma(\xx) &= g^\textit{HS}_\sigma(\etadisp(\xx)) +
  \frac{1}{4}\beta\left(\frac{\partial a_1}{\partial \etadisp(\xx)} -
  \frac{\lambdadisp}{3 \etadisp}\frac{\partial a_1}{\partial \lambdadisp}\right)
\end{align}
% This is equation 77 of gil-villegas-1997-SAFT-VR
where $g^\textit{SW}_\sigma$ is the correlation function evaluated at
contact for a hard-sphere fluid with a square-well attractive
potential, and $a_1$ and $a_2$ are two terms in the dispersive term of
the free energy.  The correlation function $g^\textit{SW}_\sigma$ is
written as a perturbative correction to the hard-sphere fluid correlation
function $g^\textit{HS}_\sigma$, for which we use the functional of Yu and
Wu\cite{yu2002fmt-dft-inhomogeneous-associating}:
\begin{align}
  g_\sigma^{HS}(\xx) &= \frac{1}{1-\etadisp(\xx)}
  +\frac32\frac{\zeta(\xx)}{(1-\etadisp(\xx))^2}
  + \frac12\frac{\etadisp(\xx)^2\zeta(\xx)}{(1-\etadisp(\xx))^3}.
\end{align}

From the Helmholtz free energy functional, we may obtain any other
thermodynamic functions we should choose.  The grand free energy
$\Phi$, which we ordinarily use in our computations, is obtained by
simply adding a chemical potential term:
\begin{equation}
  \Phi[n] = F[n] + \mu \int n(\xx) d\xx.
\end{equation}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/pressure-with-isotherms}
\end{center}
\caption{The pressure versus density for various temperatures, including
experimental pressure data from NIST\cite{nistwater}. The dash-dotted lines
indicate the computationally calculated pressure and the open circles are 
NIST data points, plotted for every other isotherm. The solid and dotted lines
represent the theoretical and experimental coexistence curves.}
\label{fig:pressure-with-isotherms}
\end{figure}

The pressure would naturally be obtained by taking a partial
derivative with respect to volume at fixed temperature.  Since
\emph{volume} isn't a parameter in density-functional theory, we
consider the \emph{free energy per unit volume}:
\begin{align}
  F[n] &= f[n]V \\
  p[n] &= -\left(\frac{\partial F[n]}{\partial V}\right)_{T} \\
  &= -\left(\frac{\partial f[n]V}{\partial V}\right)_{T} \\
  &= -V\left(\frac{\partial f[n]}{\partial V}\right)_{T}
   - f[n]\left(\frac{\partial V}{\partial V}\right)_{T} \\
  &= n \left(\frac{\partial f[n]}{\partial n}\right)_{T} - f[n].
\end{align}

Figure~\ref{fig:pressure-with-isotherms} shows the calculated pressures for
isotherms ranging from room temperature to past the critical temperature. The
solid line is the theoretical coexistence curve calculated from the SAFT
functional. For a given isotherm, the pressure curve should intersect this
coexistence curve twice, at the vapor and liquid coexistence 
pressure corresponding to the vapor and liquid densities. Even
though the theoretical pressure isotherms drop below the coexistence curve,
this is non-physical and will not actually occur. The experimental data shows
how the actual density value will ``jump'' to the other side of 
the coexistence curve at that pressure value (for
temperatures less than the critical temperature), corresponding to a phase 
change in the material. For example, follow the room temperature isotherm
starting from high pressure and high density (top right of 
Figure~\ref{fig:pressure-with-isotherms}).
As the pressure decreases, the density decreases \emph{slightly} until reaching
the coexistence curve, at which point a phase change from liquid to vapor
will occur. The
functional matches well with the experimental data from NIST away from the
critical point. The slope of the pressure versus density curve is proportional
to the compressibility, which is close but slightly too large at room
temperature.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/surface-tension}
\end{center}
\caption{The theoretical versus experimental surface tension
  versus temperature. The experimental data is taken from NIST.\cite{nistwater}
  The length-scaling parameter $\lscale$ is fit so that the theoretical surface tension
  will match the experimental surface tension near room temperature.}
\label{fig:surface-tension}
\end{figure}

@@

@@

Once we have the entropy, finding the internal energy is simple:
\begin{align}
  U[n] &= F[n] + TS[n]
\end{align}
and the enthalpy is not much more work:
\begin{align}
  H[n] &= F[n] + TS[n] + pV.
\end{align}
The calculated heat of vaporization at atmospheric pressure is
$\Delta H_{vap}$= 39.41~kJ/mol at the normal boiling point of water,
373 K. This compares well to the experimental
value measured by NIST ($\Delta H_{vap}$= 40.65~kJ/mol\cite{nistwater}), with 
an error of only a few percent.

@@

The majority of the empirical parameters used in our functional are
taken from the paper of Clark \emph{et al} on developing an optimal
SAFT model for water\cite{clark2006developing}.  This SAFT model
contains five empirical parameters: the hard-sphere radius, an energy
and length scale for the dispersion interaction, and an energy and
length scale for the association interaction.  In addition to the five
empirical parameters of Clark \emph{et al}, we add a single additional
parameter $\lscale$---with a fitted value of 0.72---which determines
the length scale over which the density is averaged when computing the
dispersion free energy.  We use this final parameter to fit the
surface tension, with the result shown in
Figure~\ref{fig:surface-tension}.  Because the SAFT model of Clark
\emph{et al} overestimates the critical temperature---which is a
common feature of models which do not explicitly treat the critical
point---we cannot correctly describe the surface tension at all
temperatures, and chose to fit it for the temperature range at which
water is liquid at one atmosphere of pressure.

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/temperature-versus-density}
\end{center}
\caption{The theoretical versus experimental coexistence curve. Away from
the critical temperature, the agreement is good. }
\label{fig:temperature-vs-density}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{figs/finding-vapor-pressure}\\
\includegraphics[width=\columnwidth]{figs/near-critical-point}
\end{center}
\caption{By plotting the free energy versus density it is possible to
  determine a saturated liquid or vapor state by the common tangent
  method. The top plot illustrates the common tangent near room 
  temperature, while the bottom plot shows the same thing close to the 
  critical temperature. }
\label{fig:homogeneous}
\end{figure}
